Covariate analysis of HBCC RNASeq data reprocessed with STAR-feature counts workflow with Ensembl v75 (DLPFC with CQN normlaisation)

### 2017-08-06 code/Rmd/HBCC_DLPFC_STAR_ENSEMBL_v75_CQN_ModSel.Rmd
Included batch effects in HBCC model selection

@@

                          "DLPFC_RNA_report: Library Batch", "DLPFC_RNA_report: Ribozero Batch", "DLPFC_RNA_report: Flowcell Batch",

@@

METADATA = METADATA %>%
  dplyr::select(Individual_ID, Institution, Gender, Age_of_Death, PMI__in_hours_, Dx, 
                DLPFC_RNA_report__Library_Batch, DLPFC_RNA_report__Ribozero_Batch, DLPFC_RNA_report__Flowcell_Batch,
                DLPFC_RNA_isolation__Exclude_, DLPFC_RNA_isolation__Sample_RNA_ID, 
                DLPFC_RNA_isolation__RIN, DLPFC_RNA_report__Exclude_, DLPFC_RNA_report__Ribozero_Batch, 
                DLPFC_RNA_report__Library_Batch, DLPFC_RNA_report__Flowcell_Batch) %>%
  dplyr::rename(LibraryBatch = DLPFC_RNA_report__Library_Batch, 
                RibozeroBatch = DLPFC_RNA_report__Ribozero_Batch, 
                FlowcellBatch = DLPFC_RNA_report__Flowcell_Batch,
                PMI = PMI__in_hours_, 
                IsolationExclude = DLPFC_RNA_isolation__Exclude_, 
                SampleID = DLPFC_RNA_isolation__Sample_RNA_ID, 
                RIN = DLPFC_RNA_isolation__RIN, 
                ReportExclude = DLPFC_RNA_report__Exclude_,
                RibozeroBatch = DLPFC_RNA_report__Ribozero_Batch, 
                LibraryBatch = DLPFC_RNA_report__Library_Batch, 
                FlowcellBatch = DLPFC_RNA_report__Flowcell_Batch) %>%
  dplyr::filter(SampleID %in% colnames(COUNT)) %>%
  dplyr::mutate(Tissue = 'DLPFC') %>%
  list(.,METADATA_QC) %>%
  join_all %>% droplevels() %>%
  unique %>% as.data.frame()
METADATA$SampleID = as.character(METADATA$SampleID)
```

@@

### Covariates clustering
Determine relationship between covariates. 
```{r covariates.clustering}
FactorCovariates <- c("Gender", "Dx.Tissue")
ContCovariates <- c("Age_of_Death", "PMI", "RIN", "TotalReads", "X5prime_Norm", "Mean_Per_Base_Cov.",
                    "Mean_CV", "Fragment_Length_StdDev", "Mapped_Pairs", "End_1_Mismatch_Rate",
                    "End_2_Mismatch_Rate", "Mapped", "IntragenicRate", "ExonicRate", "IntergenicRate",
                    "IntronicRate", "rRNARate", "GenesDetected", "TranscriptsDetected", "ExpProfEfficiency",   
                    "PercentMapped")

@@

### Library Normalisation
Initial library normalisation usign cqn
```{r cqn}
# Compute offset for gene length and gc content
CQN.GENE_EXPRESSION = cqn(NEW.COUNTS, 
                          x = GENE.GC.CONT[PROCESSED_COUNTS$filteredExprMatrix$genes$genes, 'percentage_gc_content'],
                          lengths = as.numeric(GENE.LEN[PROCESSED_COUNTS$filteredExprMatrix$genes$genes, 'Length']),
                          lengthMethod = "smooth", 
                          verbose = FALSE)
CQN.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$y + CQN.GENE_EXPRESSION$offset

# Set gene counts in specific samples that are deviating 3 sd from other samples to 0
log.mat = apply(CQN.GENE_EXPRESSION$E, 1, function(x){
  mn = mean(x, na.rm = T)
  std.dev = sd(x, na.rm = T)
  return((x < (mn-3*std.dev)) | (x > (mn+3*std.dev)))
}) %>% t
NEW.COUNTS[log.mat] = NA
CQN.GENE_EXPRESSION$E[log.mat] = NA
```

@@

adjust.covars = c(adjust.covars, vars.selected)
```
Clinical and technical variables selected in phase I are `r paste(adjust.covars, collapse = ', ')`

#### Phase II (batch effects)
```{r phase2, eval = F}
# Fit a base model with the following covariates
baseModelBIC = fitFixedEffectModel(adjust.covars, 
                                   CQN.GENE_EXPRESSION$E, 
                                   NEW.COUNTS, 
                                   COVARIATES) %>%
  dplyr::rename(baseModel = BIC)

@@

adjust.covars = c(adjust.covars, vars.selected)
```
Clinical, technical and batch variables selected after phase I and II of model identification are `r paste(adjust.covars, collapse = ', ')`

#### Phase III (RNASeq alignment specific covariates)
```{r phase3}
# Fit a base model with the following covariates
baseModelBIC = fitFixedEffectModel(adjust.covars, 
                                   CQN.GENE_EXPRESSION$E, 
                                   NEW.COUNTS, 
                                   COVARIATES) %>%
  dplyr::rename(baseModel = BIC)

@@

### Differential expression analysis
Genes that are differentially expressed at an FDR <= 0.05 are
```{r diffExp, fig.height=10, fig.width=15}
# Fit contrast
contrast = makeContrasts(contrasts=c("Dx.TissueBP_DLPFC-Dx.TissueControl_DLPFC",
                                     "Dx.TissueSCZ_DLPFC-Dx.TissueControl_DLPFC",
                                     "Dx.TissueSCZ_DLPFC-Dx.TissueBP_DLPFC"),
                         levels = colnames(ADJUSTED.FIT$coefficients))
FIT.CONTR = contrasts.fit(ADJUSTED.FIT, contrasts=contrast)
FIT.CONTR = eBayes(FIT.CONTR)

# Get differnetial expression
DE = lapply(1:3, function(i, FIT){
  topTable(FIT, coef=i, number = 50000, confint = TRUE) %>%
    rownameToFirstColumn('ensembl_gene_id') 
}, FIT.CONTR) 
names(DE) = colnames(contrast)

DE = DE %>% 
  data.table::rbindlist(idcol = 'Comparison') %>%
  dplyr::mutate(Comparison = gsub('Dx.Tissue','',Comparison),
                Direction = logFC/abs(logFC),
                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')),
                Direction = as.character(Direction)) %>%
  tidyr::separate(Comparison, into = c('ref.state','to.state'), sep = '-') %>%
  tidyr::separate(ref.state, into = c('Dx1','Tissue1'), sep = '_') %>%
  tidyr::separate(to.state, into = c('Dx2','Tissue2'), sep = '_') %>%
  left_join(Ensemble2HGNC) %>%
  left_join(GENE.LEN) %>%
  tidyr::unite(Comparison, Dx1, Dx2, sep = '_vs_')

DE$Direction[DE$adj.P.Val > 0.05 | abs(DE$logFC) < log2(1.2)] = 'NONE'

tmp = DE %>%
  dplyr::filter(adj.P.Val <= 0.05) %>%
  dplyr::select(ensembl_gene_id, Comparison, Tissue1, logFC) %>%
  group_by(Comparison, Tissue1) %>%
  dplyr::summarise(FDR_0_05_UP = length(unique(ensembl_gene_id[logFC > 0])),
                   FDR_0_05_DOWN = length(unique(ensembl_gene_id[logFC < 0])))

tmp1 = DE %>%
  dplyr::filter(adj.P.Val <= 0.05, abs(logFC) >= log2(1.2)) %>%
  dplyr::select(ensembl_gene_id, Comparison, Tissue1, logFC) %>%
  group_by(Comparison, Tissue1) %>%
  dplyr::summarise(FDR_0_05_FC_1.2_UP = length(unique(ensembl_gene_id[logFC > log2(1.2)])),
                   FDR_0_05_FC_1.2_DOWN = length(unique(ensembl_gene_id[logFC < -log2(1.2)])))

@@

p = ggplot(DE, aes(y = -log10(adj.P.Val), x = logFC, color = Direction)) + geom_point() + xlim(c(-1,1))
p = p + scale_color_manual(values = c('green','grey','red'))
p = p + facet_grid(Tissue1+.~Comparison, scales = 'fixed')
p
```

### Associate differential expression results with gc content, gene length and average expression
```{r associate.de, fig.height=10, fig.width=15}
pl = list()
pl[[1]] = ggplot(DE, aes(x = log10(as.numeric(Length)), y = logFC, color = Direction)) + geom_point() 
pl[[1]] = pl[[1]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = log10(as.numeric(Length)), y = logFC))
pl[[1]] = pl[[1]] + facet_grid(Comparison~.) + scale_color_manual(values = c('green','grey','red'))
pl[[1]] = pl[[1]] + theme(legend.position = 'top')

pl[[2]] = ggplot(DE, aes(x = percentage_gc_content, y = logFC, color = Direction)) + geom_point()
pl[[2]] = pl[[2]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = percentage_gc_content, y = logFC))
pl[[2]] = pl[[2]] + facet_grid(Comparison~.) + scale_color_manual(values = c('green','grey','red'))
pl[[2]] = pl[[2]] + theme(legend.position = 'top')

pl[[3]] = ggplot(DE, aes(x = AveExpr, y = logFC, color = Direction)) + geom_point()
pl[[3]] = pl[[3]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = AveExpr, y = logFC))
pl[[3]] = pl[[3]] + facet_grid(Comparison~.) + scale_color_manual(values = c('green','grey','red'))
pl[[3]] = pl[[3]] + theme(legend.position = 'top')

@@

### Gene set enrichment analysis
Using Fisher's exact test for gene set enrichment analysis
```{r enrich.de, fig.height=16, fig.width=16}
diffexp.genes = DE %>%
  dplyr::mutate(Direction = sign(logFC),
                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP'))) %>%
  dlply(.(Comparison, Tissue1, Direction), .fun = function(x){
    tmp = x$hgnc_symbol[x$adj.P.Val <= 0.05] %>% unique
    tmp = tmp[!is.na(tmp)]
  })
diffexp.genes = diffexp.genes[sapply(diffexp.genes, length) > 20]
backgroundGenes = unique(DE$hgnc_symbol)

@@

# Perform enrichment analysis
enrichment.results = plyr::llply(diffexp.genes, .fun = function(x, genesetsToTest, backgroundGenes){
   ldply(genesetsToTest, .fun = function(y,x,backgroundGenes){
     CovariateAnalysis::fisherEnrichment(x, y, backgroundGenes)
    }, x, backgroundGenes, .parallel = TRUE) %>%
    dplyr::mutate(fdr = p.adjust(pval, method = 'fdr'))
}, genesets.to.test, backgroundGenes) %>%
  rbindlist(idcol = 'Comparison', use.names = T, fill = T) %>%
  tidyr::separate(Comparison, c('Name', 'Region'), sep = '\\.') %>%
  dplyr::rename(Comparison = Name, Category = .id)
```

@@

# Store filtered counts
PROCESSED_COUNTS$filteredExprMatrix$counts %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  write.table(file = 'CMC_ACC_DLPFC_Counts.tsv', sep = '\t', row.names=F, quote=F)
COUNT_OBJ = File('CMC_ACC_DLPFC_Counts.tsv', name = 'Counts (filtered raw)', parentId = CODE$properties$id)
annotations(COUNT_OBJ) = annotations(CODE)
annotations(COUNT_OBJ)$dataSubType = 'filteredCounts'
COUNT_OBJ = synStore(COUNT_OBJ, activity = synGetActivity(COV_OBJ$properties$id))

>>>>

# Get RNA qc metrics from synapse
METADATA_ALG_REP_DLPFC_ID = 'syn2279446'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_ALG_REP_DLPFC_ID
METADATA_ALG_REP_DLPFC = read.table(synGet(METADATA_ALG_REP_DLPFC_ID)@filePath, 
                                    fill = T, header = T, sep = ',') %>%
  dplyr::select(Sample.RNA.ID, Ribozero.Batch, Library.Batch, Flowcell.Batch) %>%
  dplyr::rename(SampleID = Sample.RNA.ID,
                LibraryBatch = Library.Batch, 
                RibozeroBatch = Ribozero.Batch, 
                FlowcellBatch = Flowcell.Batch) %>%
  unique()

@@

@@

METADATA = METADATA %>%
  dplyr::select(Individual_ID, Institution, Gender, Age_of_Death, PMI__in_hours_, Dx, 
                DLPFC_RNA_isolation__Exclude_, DLPFC_RNA_isolation__Sample_RNA_ID, 
                DLPFC_RNA_isolation__RIN, DLPFC_RNA_report__Exclude_) %>%
  dplyr::rename(PMI = PMI__in_hours_, 
                IsolationExclude = DLPFC_RNA_isolation__Exclude_, 
                SampleID = DLPFC_RNA_isolation__Sample_RNA_ID, 
                RIN = DLPFC_RNA_isolation__RIN, 
                ReportExclude = DLPFC_RNA_report__Exclude_) %>%
  dplyr::filter(SampleID %in% colnames(COUNT)) %>%
  dplyr::mutate(Tissue = 'DLPFC') %>%
  list(.,METADATA_QC, METADATA_ALG_REP_DLPFC) %>%
  join_all %>% droplevels() %>%
  unique %>% as.data.frame()
METADATA$SampleID = as.character(METADATA$SampleID)

# Fix missing values
levels(METADATA$LibraryBatch)[1] = 'NotAvailable'
levels(METADATA$FlowcellBatch)[1] = 'NotAvailable'
levels(METADATA$RibozeroBatch)[1] = 'NotAvailable'
```

@@

### Covariates clustering
Determine relationship between covariates. 
```{r covariates.clustering}
FactorCovariates <- c("Gender", "Dx.Tissue", "LibraryBatch", "RibozeroBatch", "FlowcellBatch")
ContCovariates <- c("Age_of_Death", "PMI", "RIN", "TotalReads", "X5prime_Norm", "Mean_Per_Base_Cov.",
                    "Mean_CV", "Fragment_Length_StdDev", "Mapped_Pairs", "End_1_Mismatch_Rate",
                    "End_2_Mismatch_Rate", "Mapped", "IntragenicRate", "ExonicRate", "IntergenicRate",
                    "IntronicRate", "rRNARate", "GenesDetected", "TranscriptsDetected", "ExpProfEfficiency",   
                    "PercentMapped")

@@

### Library Normalisation
Initial library normalisation usign cqn
```{r cqn}
# Compute offset for gene length and gc content
CQN.GENE_EXPRESSION = cqn(NEW.COUNTS, 
                          x = GENE.GC.CONT[PROCESSED_COUNTS$filteredExprMatrix$genes$genes, 'percentage_gc_content'],
                          lengths = GENE.LEN[PROCESSED_COUNTS$filteredExprMatrix$genes$genes, 'Length'],
                          lengthMethod = "smooth", 
                          verbose = FALSE)
CQN.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$y + CQN.GENE_EXPRESSION$offset

# Set gene counts in specific samples that are deviating 3 sd from other samples to 3SD limit
CQN.GENE_EXPRESSION$E = apply(CQN.GENE_EXPRESSION$E, 1, function(x){
  mn = mean(x, na.rm = T)
  std.dev = sd(x, na.rm = T)

# Back calculate counts for further analysis
LIB.SIZE = colSums(PROCESSED_COUNTS$filteredExprMatrix$counts)
NEW.COUNTS = (2^CQN.GENE_EXPRESSION$E) * t(replicate(dim(CQN.GENE_EXPRESSION$E)[1], LIB.SIZE))/1e6
```

@@

adjust.covars = c(adjust.covars, vars.selected)
```
Clinical and technical variables selected in phase I are `r paste(adjust.covars, collapse = ', ')`
#### Phase II (batch effects)
```{r phase2}
# Fit a base model with the following covariates
baseModelBIC = fitFixedEffectModel(adjust.covars, 
                                   CQN.GENE_EXPRESSION$E, 
                                   NEW.COUNTS, 
                                   COVARIATES) %>%
  dplyr::rename(baseModel = BIC)

@@

adjust.covars = c(adjust.covars, vars.selected)
```
Clinical, technical and batch variables selected after phase I and II of model identification are `r paste(adjust.covars, collapse = ', ')`
#### Phase III (RNASeq alignment specific covariates)
```{r phase3}
# Fit a base model with the following covariates
baseModelBIC = fitFixedEffectModel(adjust.covars, 
                                   CQN.GENE_EXPRESSION$E, 
                                   NEW.COUNTS, 
                                   COVARIATES) %>%
  dplyr::rename(baseModel = BIC)

@@

### Differential expression analysis (SCZ - Control)
Genes that are differentially expressed at an FDR <= 0.05 are
```{r diffExp, fig.height=15, fig.width=20}
# Covariates to adjust
postAdjustCovars = adjust.covars

writeLines(paste('Using following covariates in the model:',
                 paste(postAdjustCovars, collapse=', ')))

# Post adjusted design matrix
DESIGN = getDesignMatrix(droplevels(COVARIATES[,postAdjustCovars,drop=F]),Intercept = F)$design
DESIGN = DESIGN[,linColumnFinder(DESIGN)$indepCols]

# Estimate voom weights
VOOM.GENE_EXPRESSION = limma::voom(NEW.COUNTS, design = DESIGN)
VOOM.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$E

# Fit contrast
contrast = makeContrasts(contrasts=c("Dx.TissueSCZ_DLPFC-Dx.TissueControl_DLPFC",
                                     "Dx.TissueBP_DLPFC-Dx.TissueControl_DLPFC",
                                     "Dx.TissueSCZ_DLPFC-Dx.TissueBP_DLPFC"),
                         levels = colnames(ADJUSTED.FIT$coefficients))
FIT.CONTR = contrasts.fit(ADJUSTED.FIT, contrasts=contrast)
FIT.CONTR = eBayes(FIT.CONTR)

# Get differnetial expression
DE = lapply(1:dim(contrast)[2], function(i, FIT){
  topTable(FIT, coef=i, number = 50000, confint = TRUE) %>%
    rownameToFirstColumn('ensembl_gene_id') 
}, FIT.CONTR) 
names(DE) = gsub('Dx.Tissue', '', colnames(contrast))

DE = DE %>% 
  rbindlist(idcol = 'Comparison') %>%
  dplyr::mutate(Comparison = gsub('Dx.Tissue','',Comparison),
                Direction = sign(logFC),
                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')),
                Direction = as.character(Direction)) %>%
  tidyr::separate(Comparison, into = c('to.state','ref.state'), sep = '-') %>%
  tidyr::separate(to.state, into = c('to.Dx','to.Tissue'), sep = '_') %>%
  tidyr::separate(ref.state, into = c('ref.Dx','ref.Tissue'), sep = '_') %>%
  left_join(Ensemble2HGNC) %>%
  left_join(GENE.LEN %>% dplyr::mutate(Length = as.numeric(Length)))
DE$Direction[DE$adj.P.Val > 0.05 | abs(DE$logFC) < log2(1.2)] = 'NONE'

tmp = DE %>%
  dplyr::filter(adj.P.Val <= 0.05, ref.Dx != 'Other', to.Dx != 'Other') %>%
  dplyr::select(ensembl_gene_id, to.Dx, to.Tissue, ref.Dx, ref.Tissue, logFC) %>%
  group_by(to.Dx, to.Tissue, ref.Dx, ref.Tissue) %>%
  dplyr::summarise(FDR_0_05_DOWN = length(unique(ensembl_gene_id[logFC < 0])),
                   FDR_0_05_UP = length(unique(ensembl_gene_id[logFC > 0])))

tmp1 = DE %>%
  dplyr::filter(adj.P.Val <= 0.05, abs(logFC) >= log2(1.2), ref.Dx != 'Other', to.Dx != 'Other') %>%
  dplyr::select(ensembl_gene_id, to.Dx, to.Tissue, ref.Dx, ref.Tissue, logFC) %>%
  group_by(to.Dx, to.Tissue, ref.Dx, ref.Tissue) %>%
  dplyr::summarise(FDR_0_05_FC_1.2_DOWN = length(unique(ensembl_gene_id[logFC < 0])),
                   FDR_0_05_FC_1.2_UP = length(unique(ensembl_gene_id[logFC > 0])))

@@

p = DE %>%
  dplyr::filter(ref.Dx != 'ALL', to.Dx != 'ALL') %>%
  ggplot(aes(y = -log10(adj.P.Val), x = logFC, color = Direction)) + geom_point() + xlim(c(-1,1))
p = p + scale_color_manual(values = c('green','grey','red')) + theme(legend.position = 'top')
p = p + facet_grid(ref.Dx+ref.Tissue~.+to.Dx+to.Tissue, scales = 'fixed')
p = p + ggtitle('Differential Expression between Schizophrenia and Control')
p
```

### Associate differential expression results with gc content, gene length and average expression
```{r associate.de, fig.height=10, fig.width=15}
pl = list()
pl[[1]] = ggplot(DE %>% dplyr::filter(ref.Dx == 'Control', to.Dx == 'SCZ'), 
                 aes(x = log10(Length), y = logFC, color = Direction)) + geom_point() 
pl[[1]] = pl[[1]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = log10(Length), y = logFC))
pl[[1]] = pl[[1]] + facet_grid(ref.Tissue~.) + scale_color_manual(values = c('green','grey','red'))
pl[[1]] = pl[[1]] + theme(legend.position = 'top')

pl[[2]] = ggplot(DE %>% dplyr::filter(ref.Dx == 'Control', to.Dx == 'SCZ'), 
                 aes(x = percentage_gc_content, y = logFC, color = Direction)) + geom_point()
pl[[2]] = pl[[2]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = percentage_gc_content, y = logFC))
pl[[2]] = pl[[2]] + facet_grid(ref.Tissue~.) + scale_color_manual(values = c('green','grey','red'))
pl[[2]] = pl[[2]] + theme(legend.position = 'top')

pl[[3]] = ggplot(DE %>% dplyr::filter(ref.Dx == 'Control', to.Dx == 'SCZ'), 
                 aes(x = AveExpr, y = logFC, color = Direction)) + geom_point()
pl[[3]] = pl[[3]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = AveExpr, y = logFC))
pl[[3]] = pl[[3]] + facet_grid(ref.Tissue~.) + scale_color_manual(values = c('green','grey','red'))
pl[[3]] = pl[[3]] + theme(legend.position = 'top')

@@

### Differential expression analysis (Gender specific SCZ - Control)
Genes that are differentially expressed at an FDR <= 0.05 are
```{r diffExp1, fig.height=15, fig.width=20}
# Covariates to adjust
COVARIATES$Dx.Tissue.Gender = factor(paste(COVARIATES$Dx.Tissue, COVARIATES$Gender, sep = '_'))
postAdjustCovars = c('Dx.Tissue.Gender', 'RIN', 'IntronicRate')

writeLines(paste('Using following covariates in the model:',
                 paste(postAdjustCovars, collapse=', ')))

# Post adjusted design matrix
DESIGN1 = getDesignMatrix(droplevels(COVARIATES[,postAdjustCovars,drop=F]),Intercept = F)$design
DESIGN1 = DESIGN1[,linColumnFinder(DESIGN1)$indepCols]

# Estimate voom weights
VOOM.GENE_EXPRESSION1 = limma::voom(NEW.COUNTS, design = DESIGN1)
VOOM.GENE_EXPRESSION1$E = CQN.GENE_EXPRESSION$E

# Fit linear model using new weights and new design
ADJUSTED.FIT1 = lmFit(VOOM.GENE_EXPRESSION1)

# Fit contrast
contrast = makeContrasts(contrasts=c("Dx.Tissue.GenderSCZ_DLPFC_Male-Dx.Tissue.GenderControl_DLPFC_Male",
                                     "Dx.Tissue.GenderSCZ_DLPFC_Female-Dx.Tissue.GenderControl_DLPFC_Female"),
                         levels = colnames(ADJUSTED.FIT1$coefficients))
FIT.CONTR1 = contrasts.fit(ADJUSTED.FIT1, contrasts=contrast)
FIT.CONTR1 = eBayes(FIT.CONTR1)

# Get differnetial expression
DE1 = lapply(1:dim(contrast)[2], function(i, FIT){
  topTable(FIT, coef=i, number = 50000, confint = TRUE) %>%
    rownameToFirstColumn('ensembl_gene_id') 
}, FIT.CONTR1) 
names(DE1) = gsub('Dx.Tissue.Gender', '', colnames(contrast))

DE1 = DE1 %>% 
  data.table::rbindlist(idcol = 'Comparison') %>%
  dplyr::mutate(Comparison = gsub('Dx.Tissue.Gender','',Comparison),
                Direction = sign(logFC),
                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')),
                Direction = as.character(Direction)) %>%
  tidyr::separate(Comparison, into = c('to.state','ref.state'), sep = '-') %>%
  tidyr::separate(to.state, into = c('to.Dx', 'to.Tissue', 'to.Gender'), sep = '_') %>%
  tidyr::separate(ref.state, into = c('ref.Dx', 'ref.Tissue', 'ref.Gender'), sep = '_') %>%
  left_join(Ensemble2HGNC) %>%
  left_join(GENE.LEN %>% dplyr::mutate(Length = as.numeric(Length)))
DE1$Direction[DE1$adj.P.Val > 0.05 | abs(DE1$logFC) < log2(1.2)] = 'NONE'

tmp = DE1 %>%
  dplyr::filter(adj.P.Val <= 0.05) %>%
  dplyr::select(ensembl_gene_id, to.Dx, to.Tissue, to.Gender, ref.Dx, ref.Tissue, ref.Gender, logFC) %>%
  group_by(to.Dx, to.Tissue, to.Gender, ref.Dx, ref.Tissue, ref.Gender) %>%
  dplyr::summarise(FDR_0_05_DOWN = length(unique(ensembl_gene_id[logFC < 0])),
                   FDR_0_05_UP = length(unique(ensembl_gene_id[logFC > 0])))

tmp1 = DE1 %>%
  dplyr::filter(adj.P.Val <= 0.05, abs(logFC) >= log2(1.2)) %>%
  dplyr::select(ensembl_gene_id, to.Dx, to.Tissue, to.Gender, ref.Dx, ref.Tissue, ref.Gender, logFC) %>%
  group_by(to.Dx, to.Tissue, to.Gender, ref.Dx, ref.Tissue, ref.Gender) %>%
  dplyr::summarise(FDR_0_05_FC_1.2_DOWN = length(unique(ensembl_gene_id[logFC < 0])),
                   FDR_0_05_FC_1.2_UP = length(unique(ensembl_gene_id[logFC > 0])))

p = DE1 %>%
  ggplot(aes(y = -log10(adj.P.Val), x = logFC, color = Direction)) + geom_point() + xlim(c(-1,1))
p = p + scale_color_manual(values = c('green','grey','red')) + theme(legend.position = 'top')
p = p + facet_grid(ref.Tissue~.+ref.Gender, scales = 'fixed')
p = p + ggtitle('Differential Expression between Schizophrenia and Control')
p

table(COVARIATES$Dx.Tissue, COVARIATES$Gender)
```

### Differential expression analysis (ACC - DLPFC, SCZ - Control, Female - Male)
Genes that are differentially expressed at an FDR <= 0.05 are
```{r diffExp2, fig.height=16, fig.width=8}
postAdjustCovars = c('Dx.Tissue', 'Gender', 'RIN', 'IntronicRate')

# Post adjusted design matrix
DESIGN2 = getDesignMatrix(droplevels(COVARIATES[,postAdjustCovars,drop=F]),Intercept = T)$design
DESIGN2 = DESIGN2[,linColumnFinder(DESIGN2)$indepCols]

# Estimate voom weights
VOOM.GENE_EXPRESSION2 = voom(NEW.COUNTS, design = DESIGN2)
VOOM.GENE_EXPRESSION2$E = CQN.GENE_EXPRESSION$E

# Fit linear model using new weights and new design
ADJUSTED.FIT2 = lmFit(VOOM.GENE_EXPRESSION2)

# Fit contrast
contrast = makeContrasts(contrasts=c("Dx.TissueSCZ_DLPFC-Dx.TissueControl_DLPFC",
                                     "GenderMale"),
                         levels = colnames(ADJUSTED.FIT2$coefficients))
FIT.CONTR2 = contrasts.fit(ADJUSTED.FIT2, contrasts=contrast)
FIT.CONTR2 = eBayes(FIT.CONTR2)

# Get differnetial expression
DE2 = lapply(1:dim(contrast)[2], function(i, FIT){
  topTable(FIT, coef=i, number = 50000, confint = TRUE) %>%
    rownameToFirstColumn('ensembl_gene_id') 
}, FIT.CONTR2) 
names(DE2) = c('SCZ-CONTROL', 'FEMALE-MALE')

DE2 = DE2 %>% 
  rbindlist(idcol = 'Comparison') %>%
  dplyr::mutate(Direction = sign(logFC),
                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')),
                Direction = as.character(Direction)) %>%
  left_join(Ensemble2HGNC) 
DE2$Direction[DE2$adj.P.Val > 0.05 | abs(DE2$logFC) < log2(1.2)] = 'NONE'

tmp = DE2 %>%
  dplyr::filter(adj.P.Val <= 0.05) %>%
  dplyr::select(Comparison, ensembl_gene_id, logFC) %>%
  group_by(Comparison) %>%
  dplyr::summarise(FDR_0_05_DOWN = length(unique(ensembl_gene_id[logFC < 0])),
                   FDR_0_05_UP = length(unique(ensembl_gene_id[logFC > 0])))

tmp1 = DE2 %>%
  dplyr::filter(adj.P.Val <= 0.05, abs(logFC) >= log2(1.2)) %>%
  dplyr::select(Comparison, ensembl_gene_id, logFC) %>%
  group_by(Comparison) %>%
  dplyr::summarise(FDR_0_05_FC_1.2_DOWN = length(unique(ensembl_gene_id[logFC < 0])),
                   FDR_0_05_FC_1.2_UP = length(unique(ensembl_gene_id[logFC > 0])))

p = DE2 %>%
  ggplot(aes(y = -log10(adj.P.Val), x = logFC, color = Direction)) + geom_point() + xlim(c(-1,1))
p = p + scale_color_manual(values = c('green','grey','red')) + theme(legend.position = 'top')
p = p + facet_grid(Comparison~., scales = 'free_y')
p = p + ggtitle('Differential Expression')
p
```

### Gene set enrichment analysis
Using Fisher's exact test for gene set enrichment analysis
```{r enrich.de, fig.height=16, fig.width=16}
all.diff.exp = list(DE %>%
                      tidyr::unite(Comparison1, to.Dx, to.Tissue) %>%
                      tidyr::unite(Comparison2, ref.Dx, ref.Tissue) %>%
                      tidyr::unite(Comparison, Comparison1, Comparison2, sep = '-'), 
                    DE1%>%
                      tidyr::unite(Comparison1, to.Dx, to.Tissue, to.Gender) %>%
                      tidyr::unite(Comparison2, ref.Dx, ref.Tissue, ref.Gender) %>%
                      tidyr::unite(Comparison, Comparison1, Comparison2, sep = '-'), 
                    DE2) %>%
  data.table::rbindlist(fill = TRUE, use.names = TRUE)
diffexp.genes.up = all.diff.exp %>%
  dlply(.(Comparison), .fun = function(x){
    tmp = x$hgnc_symbol[x$adj.P.Val <= 0.05 & x$logFC > 0] %>% unique
    tmp = tmp[!is.na(tmp)]
  })
names(diffexp.genes.up) = paste(names(diffexp.genes.up), 'UP', sep = '.')

diffexp.genes.down = all.diff.exp %>%
  dlply(.(Comparison), .fun = function(x){
    tmp = x$hgnc_symbol[x$adj.P.Val <= 0.05 & x$logFC < 0] %>% unique
    tmp = tmp[!is.na(tmp)]
  })
names(diffexp.genes.down) = paste(names(diffexp.genes.down), 'DOWN', sep = '.')

diffexp.genes = c(diffexp.genes.up, diffexp.genes.down)
diffexp.genes = diffexp.genes[sapply(diffexp.genes, length) > 20]
backgroundGenes = unique(DE$hgnc_symbol)

@@

# Perform enrichment analysis
enrichment.results = llply(diffexp.genes, .fun = function(x, genesetsToTest, backgroundGenes){
   ldply(genesetsToTest, .fun = function(y,x,backgroundGenes){
     CovariateAnalysis::fisherEnrichment(x, y, backgroundGenes)
    }, x, backgroundGenes, .parallel = TRUE) %>%
    dplyr::mutate(fdr = p.adjust(pval, method = 'fdr'))
}, genesets.to.test, backgroundGenes) %>%
  rbindlist(idcol = 'Comparison', use.names = T, fill = T) %>%
  dplyr::rename(Category = .id)
```

@@

# Store filtered counts
NEW.COUNTS %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  write.table(file = 'CMC_ACC_DLPFC_Counts.tsv', sep = '\t', row.names=F, quote=F)
COUNT_OBJ = File('CMC_ACC_DLPFC_Counts.tsv', name = 'Counts (filtered raw)', parentId = CODE$properties$id)
annotations(COUNT_OBJ) = annotations(CODE)
annotations(COUNT_OBJ)$dataSubType = 'filteredCounts'
COUNT_OBJ = synStore(COUNT_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
